Adaptive Preregistration for Modelling

### 2024-09-22 preregistration_modelling.qmd
Reformat callouts for consistency,  including checkboxes and italicisation. Fix typos.

Identify relevant interest groups:

@@

Briefly outline

@@

-   How will you operationalise any scenarios identified in 1.1.3? For example, how will you operationalise any qualitative changes of interest, such as ‚ 'deterioration' or 'improvement'?

-   Describe how you will evaluate and distinguish the performance of alternative scenario outcomes

-   Justify or otherwise explain how you chose these measures and determined performance criteria in relation to the analytical objectives, model purpose and modelling context, such as the risk attitudes of decision-makers and stakeholders within this system
:::

@@

List and explain critical conceptual design decisions [@Grimm:2014es], including:

-   [ ] spatial and temporal scales,

-   [ ] selection of entities and processes,

-   [ ] representation of stochasticity and heterogeneity,

-   [ ] consideration of local versus global interactions, environmental drivers, etc.

-   [ ] Explain and justify the influence of particular theories, concepts, or earlier models against alternative conceptual design decisions that might lead to alternative model structures.
:::

@@

Specify key assumptions and uncertainties underlying the model design, describing how uncertainty and variation will be represented in the model [@Moallemi2019]. Sources of uncertainty may include:

@@

Identify and define system system variables and structures, referencing scenario definitions, and the scope of the model as described within problem formulation (1.1.3):

-   [ ] What variables would support taking this action or making this decision?
-   [ ] What additional variables may interact with this system (things we can't control, but can hopefully measure)?
-   [ ] What variables have not been measured, but may interact with the system (often occurs in field or observational studies)?
-   [ ] What variables are index or surrogate measures of variables that we cannot or have not measured?
-   [ ] In what ways do we expect these variables to interact (model structures)?
-   [ ] Explain how any key concepts or terms within problem or decision-making contexts, such as regulatory terms, will be operationalised and defined in a biologically meaningful way to answer the research question appropriately? (Should this last step go here or within define conceptual framework?.
:::

@@

*This section specifies the plan for collecting, processing and preparing data available for parameterisation, determining model structure, and for scenario analysis.*
:::

@@

For each separate preliminary or investigatory analysis:

-   [ ] State what needs to be known to proceed with further decision-making about the modelling procedure, and why the analysis is necessary.

-   [ ] Explain how you will implement this analysis, as well as any techniques you will use to summarise and explore your data.

-   [ ] What method will you use to represent this analysis (graphical, tabular, or otherwise, describe)

-   [ ] Specify exactly which parts of the data will be used

-   [ ] Describe how the results will be interpreted, listing each potential analytic decision, as well as the analysis finding that will trigger each decision, where possible.
:::

@@

-   [ ] Describe how you will determine how reliable the data is for the given model purpose. Ideally, model input data should be internally consistent across temporal and spatial scales and resolutions, and appropriate to the problem at hand

-   [ ] Document any issues with data reliability.

-   [ ] How will you determine what data, if any, will be excluded from your analyses?

-   [ ] How will outliers be handled? Describe rules for identifying outlier data, and for excluding a site, transect, quadrat, year or season, species, trait, etc.

-   [ ] How will you identify and deal with incomplete or missing data?
:::

@@

### Performance Criteria

@@

Specify assumptions and key uncertainties in the formal model. Describe what gaps exist between the model conception, and the real-world problem, what biases might this introduce and how might this impact any interpretation of the model outputs, and what implications are there for evaluating model-output to inform inferences or decisions?
:::

@@

@@

Once critical decisions have been made about the modelling approach and method of model specification, the conceptual model is translated into the quantitative model.

-   [ ] Specify all formal models
    -   Note, For data-driven approaches to determining model structure and or parameterisation, it will not be able to complete this preregistration item.
-   [ ] For quantitative model selection approaches, including ensemble modelling, specify each model used in the candidate set, including any null or full/global model.
:::

@@

-   If partitioning data for cross-validation or similar approach:

@@

-   If using external / independent holdout data for model testing and evaluation:

    -   [ ] Which data will be used as a the testing data? What method will you be used for generating training / test data subsets?

    -   [ ] Describe any known differences between the training/validation and testing datasets, the relative size of each, as well as any stratification methods used for ensuring evenness of groups between data sets?

    -   [ ] It is preferable that any independent data used for model testing remains unknown to modellers during the process of model development, please describe the relationship modellers have to model validation data, will independent datasets be known or accessible to any modeller or analyst?
:::

@@

@@

-   [ ] What Quality Assurance measures will you take to verify the model has been correctly implemented? Specifying a priori quality assurance tests for implementation verification may help to avoid selective debugging and silent errors.

    -   Checks for verification implementation should include i) thoroughly checking for bugs or programming errors, and ii) whether the implemented model performs as described by the model description [@Grimm:2014es]*.*

    -   Qualitative tests could include syntax checking of code, and peer-code review [@ivimey2023]. Technical measures include using unit tests, or in-built checks within functions to prevent potential errors.
:::

@@

### Quantitative Model Checking

@@

### Qualitative Model Checking

@@

The consequences of assumption violations on the interpretation of results should be assessed [@Araujo2019].

-   Explain how you will demonstrate robustness to model assumptions and check for violations of model assumptions.

-   If you cannot perform quantitative assumption checks, describe what theoretical justifications would justify a lack of violation of or robustness to model assumptions.

-   If you cannot demonstrate or theoretically justify violation or robustness to assumptions, explain why not, and specify whether you will discuss assumption violations and their consequences for interpretation of model outputs.

-   If assumption violations cannot be avoided, explain how you will explore the consequences of assumption violations on the interpretation of results (To be completed in interim iterations of the preregistration, only if there are departures from assumptions as demonstrated in the planned tests above).
:::

@@

<!-- #NOTE, explain how model-checking is different to model validation / evaluation. What am i defining as model analysis / application?? the phases of the modelling process underlying this template need exposition somewhere -->
:::

@@

-   [ ] Please describe how you will characterise model and data uncertainties, e.g. propagating known uncertainties through the model, investigating the effect of different model scenarios with different parameters and modelling technique combinations [@Araujo2019], or empirically constructing model distributions from model output data points, and extracting summary statistics, including the mean, median, variance, and constructing confidence intervals [@Saltelli2019].

-   [ ] Relate your choice of analysis to the context and purposes of the model described in the problem formulation phase. For instance ‚discrepancies between model output and observed output may be important for forecasting models, where cost, benefit, an risk over a substantial period must be gauged, but much less critical for decision-making or management models where the user may be satisfied with knowing that the predicted ranking order of impacts of alternative scenarios or management options is likely to be correct, with only a rough indication of their sizes" [@Jakeman:2006ii].

-   [ ] Briefly describe how you will summarise the results of these in silico experiments with graphical, tabular, or other devices, such as summary statistics.

-   [ ] If the chosen modelling approach is able to explicitly articulate uncertainty due to data, measurements or baseline conditions, such as by providing estimates of uncertainty (typically in the form of probabilistic parameter covariance, [@Jakeman:2006ii]), specify which measure of uncertainty you will use.
:::

### Sensitivity Analyses

@@

-   [ ] Describe the sensitivity analysis approach you will take: deterministic sensitivity, stochastic sensitivity (variability in the model), or scenario sensitivity (effect of changes based on scenarios).

-   [ ] Describe any sensitivity analyses you will conduct by specifying which parameters will be held constant, which will be varied, and the range and intervals of values over which those parameters will be varied.

-   [ ] State the primary objective of each sensitivity analysis, for example, to identify which input variables contribute the most to model uncertainty so that these variables can be targeted for further data collection, or alternatively to identify which variables or factors contribute little to overall model outputs, and so can be 'dropped' from future iterations of the model [@Saltelli2019].
:::

@@

-   [ ] Specify any input conditions and relevant parameter values for initial environmental conditions and decision-variables under each scenario specified in section 1.

-   [ ] Describe any other relevant technical details of model application, such as methods for how you will implement any simulations or model projections.

-   [ ] What raw and transformed model outputs will you extract from the model simulations or projections, and how will you map, plot, or otherwise display and synthesise the results of scenario and model analyses.

-   [ ] Explain how you will analyse the outputs to answer your analytical objectives. For instance, describe any trade-off or robustness analyses you will undertake to help evaluate and choose between different alternatives in consultation with experts or decision-makers.
:::

>>>>

*Identify relevant interest groups:*

@@

*Briefly outline:*

@@

-  [ ] How will you operationalise any scenarios identified in 1.1.3? For example, how will you operationalise any qualitative changes of interest, such as ‚ 'deterioration' or 'improvement'?
- [ ]  Describe how you will evaluate and distinguish the performance of alternative scenario outcomes
- [ ]  Justify or otherwise explain how you chose these measures and determined performance criteria in relation to the analytical objectives, model purpose and modelling context, such as the risk attitudes of decision-makers and stakeholders within this system
:::

@@

*List and explain critical conceptual design decisions [@Grimm:2014es], including:*

-   [ ] spatial and temporal scales,
-   [ ] selection of entities and processes,
-   [ ] representation of stochasticity and heterogeneity,
-   [ ] consideration of local versus global interactions, environmental drivers, etc.
-   [ ] Explain and justify the influence of particular theories, concepts, or earlier models against alternative conceptual design decisions that might lead to alternative model structures.
:::

@@

*Specify key assumptions and uncertainties underlying the model design, describing how uncertainty and variation will be represented in the model [@Moallemi2019]. Sources of uncertainty may include:*

@@

*Identify and define system system variables and structures, referencing scenario definitions, and the scope of the model as described within problem formulation:*

-   [ ] What variables would support taking this action or making this decision?
-   [ ] What additional variables may interact with this system (things we can't control, but can hopefully measure)?
-   [ ] What variables have not been measured, but may interact with the system (often occurs in field or observational studies)?
-   [ ] What variables are index or surrogate measures of variables that we cannot or have not measured?
-   [ ] In what ways do we expect these variables to interact (model structures)?
-   [ ] Explain how any key concepts or terms within problem or decision-making contexts, such as regulatory terms, will be operationalised and defined in a biologically meaningful way to answer the research question appropriately? 
:::

@@

*This section specifies the plan for collecting, processing and preparing data available for parameterisation, determining model structure, and for scenario analysis. It also allows the researchers to disclose any prior interaction with the data.*
:::

@@

*For each separate preliminary or investigatory analysis:*

-   [ ] State what needs to be known to proceed with further decision-making about the modelling procedure, and why the analysis is necessary.
-   [ ] Explain how you will implement this analysis, as well as any techniques you will use to summarise and explore your data.
-   [ ] What method will you use to represent this analysis (graphical, tabular, or otherwise, describe)
-   [ ] Specify exactly which parts of the data will be used
-   [ ] Describe how the results will be interpreted, listing each potential analytic decision, as well as the analysis finding that will trigger each decision, where possible.
:::

@@

-   [ ] Describe how you will determine how reliable the data is for the given model purpose. Ideally, model input data should be internally consistent across temporal and spatial scales and resolutions, and appropriate to the problem at hand
-   [ ] Document any issues with data reliability.
-   [ ] How will you determine what data, if any, will be excluded from your analyses?
-   [ ] How will outliers be handled? Describe rules for identifying outlier data, and for excluding a site, transect, quadrat, year or season, species, trait, etc.
-   [ ] How will you identify and deal with incomplete or missing data?
:::

@@

### Parameter estimation / model fit performance criteria

@@

- [ ] Specify assumptions and key uncertainties in the formal model. Describe what gaps exist between the model conception, and the real-world problem, what biases might this introduce and how might this impact any interpretation of the model outputs, and what implications are there for evaluating model-output to inform inferences or decisions?
:::

@@

Once critical decisions have been made about the modelling approach and method of model specification, the conceptual model is translated into the quantitative model.
:::

@@

-   [ ] Specify all formal models
    - [ ]  Note, For data-driven approaches to determining model structure and or parameterisation, it may not be possible to respond to this preregistration item. In such cases, explain why this is the case, and how you will document the model(s) used in the final analysis.
-   [ ] For quantitative model selection approaches, including ensemble modelling, specify each model used in the candidate set, including any null or full/global model.
:::

@@

*If partitioning data for cross-validation or similar approach (delete as needed):*

@@

*If using external / independent holdout data for model testing and evaluation (delete as needed):*

-   [ ] Which data will be used as a the testing data? What method will you be used for generating training / test data subsets?
-   [ ] Describe any known differences between the training/validation and testing datasets, the relative size of each, as well as any stratification methods used for ensuring evenness of groups between data sets?
-   [ ] It is preferable that any independent data used for model testing remains unknown to modellers during the process of model development, please describe the relationship modellers have to model validation data, will independent datasets be known or accessible to any modeller or analyst?
:::

@@

::: callout-note
## Explanation & Examples

Model implementation verification is the process of ensuring that the model has been correctly implemented, and that the model performs as described by the model description [@Grimm:2014es]. This process is distinct from model checking, which assesses the model's performance in representing the system of interest [@Conn:2018hd].

-   Checks for verification implementation should include i) thoroughly checking for bugs or programming errors, and ii) whether the implemented model performs as described by the model description [@Grimm:2014es].
-   Qualitative tests could include syntax checking of code, and peer-code review [@ivimey2023]. Technical measures include using unit tests, or in-built checks within functions to prevent potential errors.
:::

@@

-   [ ] What Quality Assurance measures will you take to verify the model has been correctly implemented? Specifying a priori quality assurance tests for implementation verification may help to avoid selective debugging and silent errors.
:::

@@

### Quantitative model checking

@@

### Qualitative model checking

@@

*The consequences of assumption violations on the interpretation of results should be assessed [@Araujo2019].*

- [ ] Explain how you will demonstrate robustness to model assumptions and check for violations of model assumptions.
- [ ] If you cannot perform quantitative assumption checks, describe what theoretical justifications would justify a lack of violation of or robustness to model assumptions.
- [ ] If you cannot demonstrate or theoretically justify violation or robustness to assumptions, explain why not, and specify whether you will discuss assumption violations and their consequences for interpretation of model outputs.
- [ ] If assumption violations cannot be avoided, explain how you will explore the consequences of assumption violations on the interpretation of results (To be completed in interim iterations of the preregistration, only if there are departures from assumptions as demonstrated in the planned tests above).
:::

@@

<!-- #NOTE, explain how model-checking is different to model validation / evaluation. What am I defining as model analysis / application?? the phases of the modelling process underlying this template need exposition somewhere -->
:::

@@

- [ ] Please describe how you will characterise model and data uncertainties, e.g. propagating known uncertainties through the model, investigating the effect of different model scenarios with different parameters and modelling technique combinations [@Araujo2019], or empirically constructing model distributions from model output data points, and extracting summary statistics, including the mean, median, variance, and constructing confidence intervals [@Saltelli2019].
- [ ] Relate your choice of analysis to the context and purposes of the model described in the problem formulation phase. For instance ‚ discrepancies between model output and observed output may be important for forecasting models, where cost, benefit, an risk over a substantial period must be gauged, but much less critical for decision-making or management models where the user may be satisfied with knowing that the predicted ranking order of impacts of alternative scenarios or management options is likely to be correct, with only a rough indication of their sizes" [@Jakeman:2006ii].
- [ ] Briefly describe how you will summarise the results of these in silico experiments with graphical, tabular, or other devices, such as summary statistics.
- [ ] If the chosen modelling approach is able to explicitly articulate uncertainty due to data, measurements or baseline conditions, such as by providing estimates of uncertainty (typically in the form of probabilistic parameter covariance, [@Jakeman:2006ii]), specify which measure of uncertainty you will use.
:::

### Sensitivity analyses

@@

-   [ ] Describe the sensitivity analysis approach you will take: deterministic sensitivity, stochastic sensitivity (variability in the model), or scenario sensitivity (effect of changes based on scenarios).
-   [ ] Describe any sensitivity analyses you will conduct by specifying which parameters will be held constant, which will be varied, and the range and intervals of values over which those parameters will be varied.
-   [ ] State the primary objective of each sensitivity analysis, for example, to identify which input variables contribute the most to model uncertainty so that these variables can be targeted for further data collection, or alternatively to identify which variables or factors contribute little to overall model outputs, and so can be 'dropped' from future iterations of the model [@Saltelli2019].
:::

@@

-   [ ] Specify any input conditions and relevant parameter values for initial environmental conditions and decision-variables under each scenario specified in section 1.
-   [ ] Describe any other relevant technical details of model application, such as methods for how you will implement any simulations or model projections.
-   [ ] What raw and transformed model outputs will you extract from the model simulations or projections, and how will you map, plot, or otherwise display and synthesise the results of scenario and model analyses.
-   [ ] Explain how you will analyse the outputs to answer your analytical objectives. For instance, describe any trade-off or robustness analyses you will undertake to help evaluate and choose between different alternatives in consultation with experts or decision-makers.
:::
